Understanding the Impact of Mortgage State and Income on Household Poverty Status

### 2024-03-31 paper/paper.qmd
add ref

---
title: "Understanding the Impact of Mortgage State and Income on Household's Poverty Status"
subtitle: "An Analysis of the 2019 Revised Supplemental Poverty Measure Dataset"
author: 
  - Mary Cheng
thanks: "Code and data are available at: https://github.com/marycx/us_poverty_analysis_2019.git"
date: today
date-format: long
abstract: "This paper examines household poverty status in the United States in 2019 using the 2019 Revised Supplemental Poverty Measure (SPM) Research Dataset from the United States Census Bureau. Using logistic regression analysis, we investigate the influence of household income level and mortgage state on poverty status. Our findings reveal a higher likelihood of poverty for households' total annual earning less than $50000 and for renters compared to property owners. All households with incomes exceeding $100000 are not in poverty. These insights emphasize the need for targeted interventions to address socioeconomic inequalities and promote financial well-being among US households."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

\newpage
# Introduction

Therefore, it is important for us to analyze why some US household are in poverty in 2019 and how various demographic and financial factors may has an influence on it. Perhaps then we could understand how to reduce the number of households in poverty. We analyze the 2019 Revised SPM Research dataset from the United States Census Bureau (USCB) [@citeCes2020]. USCB is a principal agency of the U.S. Federal Statistical System, responsible for producing data about the American people and economy. We build a prediction model for households poverty status using 2019 Revised SPM dataset. This allows us to discover trends and address potential economic and societal situations that may cause the poverty status for some US households in 2019.

@@

## Data Visualization
@fig-mortgage illustrates the relationship between mortgage state and poverty status of households. It can be seen that renters constitute the highest proportion of households experiencing poverty, followed by owners without mortgages, and finally owners with mortgages. Conversely, for households who are not in poverty, owners with mortgages represent the largest segment, followed by owners without mortgages, and then renters. In addition, it can be observed that overall a larger proportion of households are not experiencing poverty. Despite variations in poverty rates across different mortgage states, the cumulative data shows economic stability among the majority of households in the United States in 2019.

```{r}
#| label: fig-mortgage
#| fig-cap: The distribution of poverty status by mortgage state
#| echo: false
#| warning: false
#| message: false
poverty2019 |>
  ggplot(aes(x = poverty_status, fill = mortgage_state)) +
  geom_bar(width = 0.2) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Poverty Status",
    y = "Number of households",
    fill = "Mortgage state"
  )
```

@@

```{r}
#| label: fig-income
#| fig-cap: The distribution of poverty status by income level
#| echo: false
#| warning: false
#| message: false
poverty2019 |>
  ggplot(aes(x = income, fill = poverty_status)) +
  geom_bar(position = "dodge") +
  coord_flip() +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Income level",
    y = "Number of households",
    fill = "Poverty status"
  )
```

\newpage
# Model
In our analysis, we utilized a Bayesian logistic regression model to examine the relationship between poverty status and two demographic factors, mortgage status and income of the household. Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

# Results
Our results are summarized in @tbl-modelresults. Our results generally matches our expectation. To avoid multicollinearity, the model excludes one variable from each category: mortgage state "Owner with mortgage" and income level "100k-150k". The intercept represents the estimated log-odds of being in poverty when all other predictors are held constant at their reference levels. In this case, the estimated log-odds of being in poverty for people who are owner with mortgage and their total household income level to be 100k-150k annually is −6.782.

@@

# Discussion

>>>>

---
title: "Understanding the Impact of Mortgage State and Income on Household Poverty Status"
subtitle: "An Analysis of Household Poverty Status in the United States in 2019"
author: 
  - Mary Cheng
thanks: "Code and data are available at: https://github.com/marycx/us_poverty_analysis_2019.git"
date: today
date-format: long
abstract: "This paper examines household poverty status in the United States in 2019 using the 2019 Revised Supplemental Poverty Measure (SPM) Research Dataset from the United States Census Bureau. Using logistic regression analysis, we investigate the influence of household income level and mortgage state on poverty status. Our findings reveal a higher likelihood of poverty for households' total annual earning less than $50000 and for renters compared to property owners. All households with incomes exceeding $100000 are not in poverty. These insights emphasize the need for targeted interventions to address socioeconomic inequalities and promote financial well-being among US households."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

\newpage
# Introduction
Over the last few decades, addressing poverty has remained a persistent issue in the United States, leading to the implementation of multiple policies in an effort to mitigate this problem [@Kearney_Harris]. In 2019, the United States experienced the culmination of an economic expansion that followed the Great Recession, marking the last full year before the onset of COVID-19 pandemic. The robustness of the U.S. economy during this period was evidenced by low poverty rates, indicating a notable decrease in economic hardship characterized by low income. The poverty rate in the United States was estimated to be 10.5% in 2019, down from 11.8% in 2018, marking the fifth consecutive year of decline [@Semega_2022]. Despite various interventions and economic policies, understanding the intricate dynamics influencing household poverty status is still difficult. Therefore, it is important for us to analyze why some US household are in poverty in 2019 and how various demographic and financial factors may has an influence on it. Perhaps then we could understand how to reduce the number of households in poverty. We analyze the 2019 Revised SPM Research dataset from the United States Census Bureau (USCB) [@Bureau_2022]. USCB is an agency of the U.S. Federal Statistical System, responsible for producing data about the American people and economy [@uscb]. We build a prediction model for households poverty status using 2019 Revised SPM dataset. This allows us to discover trends and address potential economic and societal situations that may cause the poverty status for some US households in 2019.

@@

## Data Visualization
@fig-mortgage illustrates the relationship between mortgage state and poverty status of households. It can be seen that renters are the highest proportion of households experiencing poverty, followed by owners without mortgages, and finally owners with mortgages. Conversely, for households who are not in poverty, owners with mortgages represent the largest segment, followed by owners without mortgages, and then renters. In addition, it can be observed that overall a larger proportion of households are not experiencing poverty. Despite variations in poverty rates across different mortgage states, the cumulative data shows economic stability among the majority of households in the United States in 2019.

```{r}
#| label: fig-mortgage
#| fig-cap: The distribution of poverty status by mortgage state
#| echo: false
#| warning: false
#| message: false
poverty2019 |>
  ggplot(aes(x = mortgage_state, fill = poverty_status)) +
  geom_bar(position = "dodge", width = 0.5) +
  theme_minimal() +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Mortgage state",
    y = "Number of households",
    fill = "Poverty Status"
  )
```

#```{r}
##| label: fig-mortgage
##| fig-cap: The distribution of poverty status by mortgage state
##| echo: false
##| warning: false
##| message: false
#poverty2019 |>
#  ggplot(aes(x = poverty_status, fill = mortgage_state)) +
#  geom_bar(width = 0.2) +
#  theme_minimal() +
#  scale_fill_brewer(palette = "Set1") +
#  labs(
#    x = "Poverty Status",
#    y = "Number of households",
#    fill = "Mortgage state"
#  )
#```

@@

```{r}
#| label: fig-income
#| fig-cap: The distribution of poverty status by income level
#| echo: false
#| warning: false
#| message: false
poverty2019 |>
  ggplot(aes(x = income, fill = poverty_status)) +
  geom_bar(position = "dodge", width = 0.5) +
  coord_flip() +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Income level",
    y = "Number of households",
    fill = "Poverty status"
  )
```

# Model {#sec-model}
In our analysis, we utilized a Bayesian logistic regression model to examine the relationship between poverty status and two demographic factors, mortgage status and income of the household. Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

# Results {#sec-result}
Our results are summarized in @tbl-modelresults. Our results generally matches our expectation. To avoid multicollinearity, the model excludes one variable from each category: mortgage state "Owner with mortgage" and income level "100k-150k". The intercept represents the estimated log-odds of being in poverty when all other predictors are held constant at their reference levels. In this case, the estimated log-odds of being in poverty for people who are owner with mortgage and their total household income level to be 100k-150k annually is −6.782.

@@

# Discussion {#sec-discussion}
